### 2017-02-08 manuscript.rmd
re knit from scratch

@@

Phylogenetic comparative methods account for evolutionary history and explicitly model trait change along the branches of evolutionary trees (*e.g.*, Figure 1b). The value of these methods relative to to pairwise comparisons has been repeatedly shown in analyses of other types of character data [@Ricklefs:1996wb; @Chamberlain:2012vx; @OMeara:2012gq]. The application of these methods is illustrated by a study of leaf functional attributes for about 100 plant species [@Ackerly:1999iu]. If phylogenetic relationships are not considered, analyses indicate a negative correlation of leaf lifespan with leaf size. With the application of phylogenetic comparative methods, however, this correlation disappears. This is because most of the observed variance is due to differences between conifers and flowering plants. Phylogenetic comparative methods capture this shift as change along a single branch deep in the tree, revealing that there is not a tendency of correlated change between these traits across the phylogeny. One reason that comparative functional genomics has not yet embraced phylogenetic approaches is that it has not yet been concretely demonstrated that pairwise and phylogenetic comparative methods can lead to different results when considering functional genomics data. Here we examine this issue be re-evaluating the pairwise comparisons used in two recent studies that compared gene expression across species.

@@

@@

```{r show_contrast_result, echo=FALSE, cache=TRUE, message=F, warning=F, fig.height=2, fig.width=4, fig.cap="Phylogneetic reanalysis of the Tau values calculated by KMRR [@KryuchkovaMostacci:2016iw]. Density plot of the magnitude of phylogenetic independent contrasts for Tau following duplication and speciation events. As the contrasts for duplication events are not larger than those of speciation events, the results are not consistent with the predictions of the ortholog conjecture."}

@@

We did not find increased evolutionary change in expression following duplication events (Figure 3). The Wilcoxon rank test does not reject the null hypothesis that the rate of evolution following duplications is the same as or less than the rate following speciation (p value = `r wilcox_test_result`). Our phylogenetic comparative analysis, unlike the previously published pairwise comparative analysis [@KryuchkovaMostacci:2016iw], therefore find no evidence for the ortholog conjecture in this system.

@@

```{r pairwise, echo=FALSE, cache=TRUE, message=F, warning=F, fig.height=2, fig.width=5, fig.cap="The pairwise analyses presented in Figure 2a of KMRR [@KryuchkovaMostacci:2016iw] can be reproduced with the subset of data we consider here. Each circle indicates the Tau Pearson correlation coefficient for a set of pairwise comparisons annotated with a specific node name of a given age and event type, *i.e.*, whether the divergence at that nodes was due to speciation (giving rise to orthologs) or duplication (giving rise to paralogs). Following KMRR Figure 2a, this figure shows only the comparisons that include a human sequence."}
	get_pairwise_summary = function ( nhx ) {

@@

In order to better understand why our phylogenetic analysis supports a different conclusion (*i.e.*, no support for the ortholog conjecture) than the published analysis of KMRR [@KryuchkovaMostacci:2016iw] (*i.e.*, strong support for the ortholog conjecture), we first checked to make sure we could reproduce their result based on pairwise analyses. This is important since we are only looking at at a subset of the data they considered, the Brawand *et al.* 2011 [@Brawand:2011du] dataset for gene trees that could be successfully time calibrated. In their Figure 1, they present a higher Tau correlation coefficient between ortholog pairs than paralog ortholog pairs. We find the same here, with correlation coefficients of `r round( ortholog_r, 3)` for orthologs and `r round( paralog_r, 3)` for paralogs.

Why is it that pairwise methods and phylogenetic methods lead to opposite conclusions? One reason is that multiple pairwise comparisons repeatedly sample the same evolutionary changes and in so doing violate statistical assumptions of independence, whereas phylogenetic comparative methods make multiple independent comparisons across non-overlapping branches of the tree.  The other reason for the different conclusions is that pairwise comparisons between the tips and phylogenetic comparative methods describe different things. Pairwise comparisons describe contemporary patterns, while phylogenetic methods infer historical processes [@OMeara:2012gq]. There need not be a different process of evolution following speciation and duplication for paralogs to be more different than orthologs. This could be due to the structure of the gene phylogenies. If paralogs tend to be more distantly related to each other than orthologs, then there would be more time for differences to accumulate even if the rate of change is the same between the two. This is, in fact, the case for these data. While the mean distance between orthologs is `r round( ortholog_distance_mean, 1 )` million years, the mean distance between paralogs is `r round( paralog_distance_mean, 1 )` million years. This is because the oldest speciation event is by definition the most recent common ancestor of the species included in the study, but many gene families underwent duplication before this time.

To test the hypothesis that distantly paralogs that arose from duplications that precede the oldest speciation event (Supplementary Figure 8a) impact the lower correlation of Tau between paralogs than between orthologs, we removed them. When we consider only the duplication events the same age or younger than the oldest speciation event (Supplementary Figure 8b), the paralog correlation coefficient increases from `r round( paralog_r, 3)` to  `r round( paralog_subset_r, 3)`. This is much closer to the ortholog correlation of `r round( ortholog_r, 3) `.

@@

```{r simulations_plot, cache=TRUE, echo=FALSE, warning=FALSE, message=FALSE, fig.cap="Analyses of simulated data reveal that pairwise methods produce similar results under both the null model and the ortholog conjecture (OC), while the phylogenetic method provides different results under each model. (a) Pairwise analysis of dataset simulated under the null model that there is no difference in the rate of expresison evolution following duplication and speciation. (b) Phylogenetic analysis of data simulated under the null model. (c) Pairwise analysis of dataset simulated under the ortholog conjecture, in which there is a higher rate of expresison evolution following duplication than speciation. This plot is very similar to the corresponding plot for the null model. (d) Phylogenetic analysis of data simulated under the ortholog conjecture. This plot is distinct from that for the null model in that phylogenetic independent contrasts for duplication events are significantly larger than after speciation events."}

@@

Our results suggest, at a minimum, that the ortholog conjecture is not a dominant pattern that is central to explaining the evolution of phenotypic diversity in gene families. The ortholog conjecture does not have to be an all or nothing question, though. It may be the case that the rates of phenotypic evolution following duplication may be greater than that following duplication in some organisms, gene families, and evolutionary processes [@Gabaldon:2013fi]. We just don't find evidence for it when summarizing gene expression across tissues with Tau in these organisms. This calls into question the general predictive power of the ortholog conjecture, and until these processes are better understood it will be necessary to test for it in each situation. These tests should be articulated in terms of clear alternative hypotheses [@Studer:2009bm] that make distinct phylogenetic comparative predictions.

@@

For each node in the remaining calibrated trees, we calculated the phylogenetic independent contrast for Tau across its daughter branches with the `pic()` function in ape [@Paradis:2004dv]. We then collected the contrasts from all trees into a single table, along with otherr annotations including whether the node is a speciation or duplication event. This table, `nodes_contrast`, was then analyzed as described in the main text for the presented plots and tests.

@@

Thanks to Steve Haddock, Alex Damian Serrano, and August Guang for feedback on the manuscript. This work was supported by the National Science Foundation (DEB-1256695 and the Waterman Award).

>>>>

	#' Summarize pairwise comparisons between tips of a tree
	#' 
	#' @param nhx A phylogenetic tree as a treeio::treedata object 
	#' 	
	#' @return A dataframe
	#' @export
	get_pairwise_summary = function ( nhx ) {

@@

Phylogenetic comparative methods account for evolutionary history and explicitly model trait change along the branches of evolutionary trees (*e.g.*, Figure 1b). The value of these methods relative to to pairwise comparisons has been repeatedly shown in analyses of other types of character data [@Ricklefs:1996wb; @Chamberlain:2012vx; @OMeara:2012gq]. The application of these methods is illustrated by a study of leaf functional attributes for about 100 plant species [@Ackerly:1999iu]. If phylogenetic relationships are not considered, analyses indicate a negative correlation of leaf lifespan with leaf size. With the application of phylogenetic comparative methods, however, this correlation disappears. This is because most of the observed variance is due to differences between conifers and flowering plants. Phylogenetic comparative methods capture this shift as change along a single branch deep in the tree, revealing that there is not a tendency of correlated change between these traits across the phylogeny. One reason that comparative functional genomics has not yet embraced phylogenetic approaches is that it has not yet been concretely demonstrated that pairwise and phylogenetic comparative methods can lead to different results when considering functional genomics data. Here we examine this issue by re-evaluating the pairwise comparisons used in two recent studies that compared gene expression across species.

@@

	# Remove the large string objects to free up memory
	rm( lines )
	rm( tree_lines )

@@

```{r show_contrast_result, echo=FALSE, cache=TRUE, message=F, warning=F, fig.height=2, fig.width=4, fig.cap="Phylogenetic reanalysis of the Tau values calculated by KMRR [@KryuchkovaMostacci:2016iw]. Density plot of the magnitude of phylogenetic independent contrasts for Tau following duplication and speciation events. As the contrasts for duplication events are not larger than those of speciation events, the results are not consistent with the predictions of the ortholog conjecture."}

@@

We did not find increased evolutionary change in expression following duplication events (Figure 3). The Wilcoxon rank test does not reject the null hypothesis that the rate of evolution following duplications is the same as or less than the rate following speciation (p value = `r wilcox_test_result`). Our phylogenetic comparative analysis, unlike the previously published pairwise comparative analysis [@KryuchkovaMostacci:2016iw], therefore finds no evidence for the ortholog conjecture in this system.

@@

```{r pairwise, echo=FALSE, cache=TRUE, message=F, warning=F, fig.height=2, fig.width=5, fig.cap="The pairwise analyses presented in Figure 2a of KMRR [@KryuchkovaMostacci:2016iw] can be reproduced with the subset of data we consider here. Each circle indicates the Tau Pearson correlation coefficient for a set of pairwise comparisons annotated with a specific node name of a given age and event type, *i.e.*, whether the divergence at that nodes was due to speciation (giving rise to orthologs) or duplication (giving rise to paralogs). Following KMRR Figure 2a, this figure shows only the comparisons that include a human sequence."}

@@

In order to better understand why our phylogenetic analysis supports a different conclusion (*i.e.*, no support for the ortholog conjecture) than the published analysis of KMRR [@KryuchkovaMostacci:2016iw] (*i.e.*, strong support for the ortholog conjecture), we first checked to make sure we could reproduce their result based on pairwise analyses. This is important since we are only looking at at a subset of the data they considered, the Brawand *et al.* 2011 [@Brawand:2011du] dataset for gene trees that could be successfully time calibrated. In their Figure 1, they present a higher Tau correlation coefficient between ortholog pairs than paralog ortholog pairs. We find the same here, with correlation coefficients of `r round( ortholog_r, 2)` for orthologs and `r round( paralog_r, 2)` for paralogs.

Why is it that pairwise methods and phylogenetic methods lead to opposite conclusions? One reason is that multiple pairwise comparisons repeatedly sample the same evolutionary changes and in so doing violate statistical assumptions of independence, whereas phylogenetic comparative methods make multiple independent comparisons across non-overlapping branches of the tree. The other reason for the different conclusions is that pairwise comparisons between the tips and phylogenetic comparative methods describe different things. Pairwise comparisons describe contemporary patterns, while phylogenetic methods infer historical processes [@OMeara:2012gq]. There need not be a different process of evolution following speciation and duplication for paralogs to be more different than orthologs. This could be due to the structure of the gene phylogenies. If paralogs tend to be more distantly related to each other than orthologs, then there would be more time for differences to accumulate even if the rate of change is the same between the two. This is, in fact, the case for these data. While the mean distance (*i.e.*, total branch length) between orthologs is `r round( ortholog_distance_mean, 1 )` million years, the mean distance between paralogs is `r round( paralog_distance_mean, 1 )` million years. This is because the oldest speciation event is by definition the most recent common ancestor of the species included in the study, but many gene families underwent duplication before this time.

To test the hypothesis that ancient duplications that precede the oldest speciation event (Supplementary Figure 8a) impact the lower correlation of Tau between paralogs than between orthologs, we removed them. When we consider only the duplication events the same age or younger than the oldest speciation event (Supplementary Figure 8b), the paralog correlation coefficient increases from `r round( paralog_r, 2)` to `r round( paralog_subset_r, 2)`. This is much closer to the ortholog correlation of `r round( ortholog_r, 2) `.

@@

```{r simulations_plot, cache=TRUE, echo=FALSE, warning=FALSE, message=FALSE, fig.cap="Analyses of simulated data reveal that pairwise methods produce similar results under both the null model and the ortholog conjecture (OC), while the phylogenetic method provides different results under each model. (a) Pairwise analysis of dataset simulated under the null model that there is no difference in the rate of expression evolution following duplication and speciation. (b) Phylogenetic analysis of data simulated under the null model. (c) Pairwise analysis of dataset simulated under the ortholog conjecture, in which there is a higher rate of expression evolution following duplication than speciation. This plot is very similar to the corresponding plot for the null model. (d) Phylogenetic analysis of data simulated under the ortholog conjecture. This plot is distinct from that for the null model in that phylogenetic independent contrasts for duplication events are significantly larger than after speciation events."}

@@

Our results suggest, at a minimum, that the ortholog conjecture is not a dominant pattern that is central to explaining the evolution of phenotypic diversity in gene families. The ortholog conjecture does not have to be an all or nothing question, though. It may be the case that the rates of phenotypic evolution following duplication may be greater than that following duplication in some organisms, gene families, and evolutionary processes [@Gabaldon:2013fi]. We just don't find evidence for it when summarizing gene expression across tissues with Tau in these organisms. This calls into question the general predictive power of the ortholog conjecture with respect to gene expression, and until these processes are better understood it will be necessary to test for it in each situation. These tests should be articulated in terms of clear alternative hypotheses [@Studer:2009bm] that make distinct phylogenetic comparative predictions.

@@

For each node in the remaining calibrated trees, we calculated the phylogenetic independent contrast for Tau across its daughter branches with the `pic()` function in ape [@Paradis:2004dv]. We then collected the contrasts from all trees into a single table, along with other annotations including whether the node is a speciation or duplication event. This table, `nodes_contrast`, was then analyzed as described in the main text for the presented plots and tests.

@@

Thanks to Steve Haddock, Alex Damian Serrano, August Guang, and Zack Lewis for feedback on the manuscript. This work was supported by the National Science Foundation (DEB-1256695 and the Waterman Award).
